Arrows for Parallel Computations

### 2017-09-12 paper/content/motivation.tex
Motivation.tex remarks Phil

Pure functional programming languages have been used
since a longer time as a ``sand box'' for testing novel parallel
programming approaches. Reasons for this include referential
transparency and the ease of introducing new sophisticated language
concepts. Specially for Haskell (on which this paper focuses), various
approaches reached maturity. The definite parallel Haskell does not
exist, there are multiple parallel Haskells \mbcomment{Wortwiederholung? Oder okay so?}. We regard here in-depth
the Multicore Haskell (a SMP implementation of GpH language), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. 
%In this paper we introduce a notion of parallel
%computations using Arrows.
Here, we use Arrows to represent parallel computations. This has not
been done before.

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and various implementations.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface and varying implementations in
existing parallel Haskells: Multicore Haskell, |Par| Monad, and
Eden. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel Haskell backends at will. Further backends are viable, too.

\paragraph{Contribution}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
% why do we aim to abstract parallelism this way and what does this
% approach do better than the other parallel Haskells?
 is such a parallelism abstraction of benefit and to what extent does
 it improve existing approaches?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator. The resulting Arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
\end{itemize}

>>>>

One particular reason for this is the ease of introducing new sophisticated language
concepts. Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms including the expression of parallelism. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
the Multicore Haskell (a SMP implementation of GpH language), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. 
%In this paper we introduce a notion of parallel
%computations using Arrows.
A key novelty here is to use Arrows to represent parallel computations. They seem a natural fit as they are a generalization of the function |->| and serve as general interface to computations.

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and implementations for three parallel Haskells.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface and varying implementations in
existing parallel Haskells: Multicore Haskell, |Par| Monad, and
Eden. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
the parallel Haskell backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.

\paragraph{Contributions}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
% why do we aim to abstract parallelism this way and what does this
% approach do better than the other parallel Haskells?
 is such a parallelism abstraction of benefit and to what extent does
 it improve existing approaches?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator. The resulting Arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
\end{itemize}
